Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-10-05 paper.tex
to the end of L2 proof -- @lenacoll please check this proof!

In line with the research programme proposed in \autocite{Gavryushkin2018-ol}, we investigate the geometry and algorithmic complexity of the $\rnni$ space.
Specifically, in this paper we establish the exact radius and diameter of the space, [... -- Lena, please finish this including the approaximate algorithm that works exactly on small spaces (Lena: is it already in the text?)]
The question of whether there exists a polynomial algorithms for computing the $\rnni$ distance remain an open problem.

@@

@@

\begin{lemma}
$\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(\rnni)$ denotes the diameter of the $\rnni$ graph, that is, if $d(u, v)$ is the graph distance in $\rnni$ then $\Delta(\rnni) = \max \{d(u, v) \mid u, v \in \rnni\}$.
\label{lemma:diameter_bound}
\end{lemma}

@@

Lemma~\ref{lemma:distance_delete_taxon} is important for some of the following proofs using induction.

In most of the following inductive proofs the length of a shortest paths $p$ on trees on $n+1$ taxa is compared with a path on $n$ taxa induced by $p$.
The following lemma bounds the length of these induced paths utilising the rank differences of certain internal nodes between start and end tree of $p$.
Henceforth the notion $\parent(x)$ refers to the internal node that is parent of taxon $x$.

\begin{lemma}
    Let $T$ and $R$ be two trees on $n+1$ taxa.
    Let $T{\big|}_n$ and $R{\big|}_n$ be these trees restricted to taxa $1, \ldots, n$, i.e. taxon $n+1$ is deleted from $T$ and $R$ and the thereby arising node of degree two is suppressed.
    Then it is $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - \delta$ where $\delta:= |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$ is the rank difference of the parent of taxon $x$ between trees $T$ and $R$.
    \label{lemma:distance_delete_taxon}
\end{lemma}

\begin{proof}
    Let $p$ be a shortest path from tree $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
    This new path $p{\big|}_n$ is then a path from $T$ to $R$.
    Showing that at least $\delta$ $\rnni$ moves are deleted from $p$ to receive $p{\big|}_n$ results in $d(T{\big|}_n,R{\big|}_n) \leq |p{\big|}_n| \leq |p| - \delta = d(T,R) -\delta$ which proves the lemma.
    By the following two observations one can easily see that this claim is true, indeed:
    The rank of an internal node $v$ only changes within an $\rnni$ move if this move is a rank move of $v$ or an $\nni$ move on an edge incident to $v$.
    Furthermore, any $\rnni$ move can change the rank of an internal node by at most one.
    Applying this to $\parent(n+1)$ and knowing that the rank of this node differs by $\delta$ between $T$ and $R$ proves that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - \delta$.

    % Alternative: Induction:
    %
    % Induction on $k = d(T,R)$:
    %
    % Basis($k = 1$):
    % If rank of $\parent(n+1)$ changes between $T$ and $R$, it is $d(T{\big|}_n, R{\big|}_n) = 0$ and $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = 1$ and therefore, $d(T{\big|}_n, R{\big|}_n) = 0 \leq 1 - 1 = d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
    % If on the other hand the rank of $\parent(n+1}$ does not change between $T$ and $R$, it is $d(T{\big|}_n, R{\big|}_n) = d(T,R) = 1$ and $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = 1$ and therefore, $d(T{\big|}_n, R{\big|}_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
    %
    % Step($k \to k+1$):
    % Let $T$ and $R$ be trees on $n+1$ taxa with distance $d(T,R) = k$.
    % Let $R'$ be the last tree on a shortest path from $T$ to $R$, i.e. $d(T,R') = k$.
    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T{\big|}_n, R'{\big|}_n) \leq d(T, R') - |r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| := k - m$.
    % Now we differ two cases:
    % \begin{enumerate}
    %     \item $\parent(n+1}$ changes its rank between $R$ and $R'$ or is affected by the $\nni$ move is performed between $R$ and $R'$.
    %
    %     Then the trees $R{\big|}_n$ and $R'{\big|}_n$ are equal. \todo{Is this obivous?}
    %     Therefore, it is $d(T{\big|}_n,R{\big|}_n) = d(T{\big|}_n, R'{\big|}_n) \leq k - m$.
    %     It is either $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m+1$, $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m$, or $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m-1$.
    %     In either case, we have $d(T{\big|}_n,R{\big|}_n) = d(T{\big|}_n, R'{\big|}_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
    %
    %     \item $\parent(n+1}$ has same rank in $R$ and $R'$.
    %
    %     In this case it is $d(T{\big|}_n,R{\big|}_n) \leq d(T{\big|}_n,R'{\big|}_n) + d(R'{\big|}_n, R{\big|}_n) \leq d(T{\big|}_n, R'{\big|}_n) + 1 \leq (k-m)+ 1$.
    %     It also follows that $|r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| = |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$ and therefore, we can conclude $d(T{\big|}_n,R{\big|}_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
    % \end{enumerate}
\end{proof}

>>>>

In line with the research programme proposed in \autocite{Gavryushkin2018-ol}, we investigate the geometry and algorithmic complexity of the $\rnni$ space.
Specifically, in this paper we establish the exact radius and diameter of the space, [... -- Lena, please finish this including the approaximate algorithm that works exactly on small spaces (Lena: is it already in the text?)]
The question of whether there exists a polynomial algorithm for computing the $\rnni$ distance remains an open problem.

@@

We use the notation $d(T, R)$ throughout the paper to denote the graph distance between trees in $\rnni$, that is, $d(T, R)$ is the length of a shortest $\rnni$ path between $T$ and $R$.

@@

\todo{make sure trees are consistently denoted by upper case letters}
\begin{lemma}
$\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(\rnni)$ denotes the diameter of the $\rnni$ graph, that is, if $d(T, R)$ is the graph distance in $\rnni$ then $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$.
\label{lemma:diameter_bound}
\end{lemma}

@@

The following lemma relates distances between trees on $n$ and $n+1$ taxa and is an important tool for inductive arguments.
We use the notion $\parent_T(x)$ to refer to the node adjacent to taxon $x$ in tree $T$.
$T{\big|}_n$ denotes the restriction of tree $T$ to the set of taxa $\{1, \ldots, n\}$.
In particular if $T$ is a tree on taxa $1, \ldots, n+1$ the tree $T{\big|}_n$ is obtained by deleting taxon $n+1$ and suppressing the thereby created node of degree two.

\begin{lemma}
Let $T$ and $R$ be two trees on taxa $\{1, \ldots, n+1\}$.
Then $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - \delta$, where $\delta = |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
\label{lemma:distance_delete_taxon}
\end{lemma}

\begin{proof}
First observe that the rank of an internal node $v$ of a tree $T$ along an $\rnni$ path $p$ can only be changed by performing a rank move that involves $v$ or an $\nni$ move on an edge adjacent to $v$.
Second observe that any $\rnni$ move can change the rank of an internal node by at most one.

Let $p$ be a shortest path from tree $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
This new path $p{\big|}_n$ is then a path from $T{\big|}_n$ to $R{\big|}_n$.
Recall that $\delta$ is the difference in ranks between the parents of taxon $n+1$ in $T$ and $R$.
\todo{This statement has to be checked very carefully and one or two sentences added to explain the `applying'.}
Applying the two observations above we conclude that $|p{\big|}_n| \leq |p| - \delta$.
Since $d(T{\big|}_n,R{\big|}_n) \leq |p{\big|}_n|$ and $|p| = d(T,R)$, the desired inequality follows.
\end{proof}

    % Alternative: Induction:
    %
    % Induction on $k = d(T,R)$:
    %
    % Basis($k = 1$):
    % If rank of $\parent(n+1)$ changes between $T$ and $R$, it is $d(T{\big|}_n, R{\big|}_n) = 0$ and $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = 1$ and therefore, $d(T{\big|}_n, R{\big|}_n) = 0 \leq 1 - 1 = d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
    % If on the other hand the rank of $\parent(n+1}$ does not change between $T$ and $R$, it is $d(T{\big|}_n, R{\big|}_n) = d(T,R) = 1$ and $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = 1$ and therefore, $d(T{\big|}_n, R{\big|}_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
    %
    % Step($k \to k+1$):
    % Let $T$ and $R$ be trees on $n+1$ taxa with distance $d(T,R) = k$.
    % Let $R'$ be the last tree on a shortest path from $T$ to $R$, i.e. $d(T,R') = k$.
    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T{\big|}_n, R'{\big|}_n) \leq d(T, R') - |r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| := k - m$.
    % Now we differ two cases:
    % \begin{enumerate}
    %     \item $\parent(n+1}$ changes its rank between $R$ and $R'$ or is affected by the $\nni$ move is performed between $R$ and $R'$.
    %
    %     Then the trees $R{\big|}_n$ and $R'{\big|}_n$ are equal. \todo{Is this obivous?}
    %     Therefore, it is $d(T{\big|}_n,R{\big|}_n) = d(T{\big|}_n, R'{\big|}_n) \leq k - m$.
    %     It is either $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m+1$, $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m$, or $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m-1$.
    %     In either case, we have $d(T{\big|}_n,R{\big|}_n) = d(T{\big|}_n, R'{\big|}_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
    %
    %     \item $\parent(n+1}$ has same rank in $R$ and $R'$.
    %
    %     In this case it is $d(T{\big|}_n,R{\big|}_n) \leq d(T{\big|}_n,R'{\big|}_n) + d(R'{\big|}_n, R{\big|}_n) \leq d(T{\big|}_n, R'{\big|}_n) + 1 \leq (k-m)+ 1$.
    %     It also follows that $|r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| = |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$ and therefore, we can conclude $d(T{\big|}_n,R{\big|}_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
    % \end{enumerate}
